# Lab 3: R with Positron (Solutions)

### Setup
```{r, results='hide', message=FALSE}
install.packages(c("tidyverse", "ggplot2", "bestNormalize", "ARTool", "lmerTest", "xtable"))
library(tidyverse)
library(ggplot2)
library(bestNormalize)
library(ARTool)
library(lmerTest)
library(xtable)
```

## Section 1: Data Loading and Inspection

### 1.1

```{r}
# setwd("path/to/GREEN-LAB/Lab 3")
dat_data <- read.csv("data/Run_Table_TPCH.csv") %>%
  select(library, dataframe_size, trial, energy_usage, execution_time, cpu_usage, memory_usage) %>%
  mutate(
    library = factor(library),
    dataframe_size = factor(dataframe_size),
    trial = factor(trial)
  )

glimpse(dat_data)
summary(dat_data)
head(dat_data)
```

## Section 2: Visual Exploration via Plotting

### 2.1

```{r}
# Histogram of overall energy usage
ggplot(dat_data, aes(x = energy_usage)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Histogram of Energy Usage",
    x = "Energy Usage",
    y = "Frequency"
  )

# Histogram split by library
ggplot(dat_data, aes(x = energy_usage, fill = library)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  theme_minimal() +
  labs(
    title = "Energy Usage Distribution by Library",
    x = "Energy Usage",
    y = "Frequency",
    fill = "Library"
  )
```

### 2.2

```{r}
ggplot(dat_data, aes(x = library, y = energy_usage, fill = dataframe_size)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Energy Usage by Library and Dataframe Size",
    x = "Library",
    y = "Energy Usage",
    fill = "Dataframe Size"
  )
```

### 2.3

```{r}
# Define "experiment" as library + dataframe_size
dat_data <- dat_data %>%
  mutate(experiment = interaction(library, dataframe_size))

ggplot(dat_data, aes(x = trial, y = energy_usage, color = experiment)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal() +
  labs(
    title = "Energy Usage Across Trials",
    x = "Trial",
    y = "Energy Usage",
    color = "Experiment"
  )
```

## Section 3: Data Normality and Normalization

### 3.1

```{r}
# Run Shapiro-Wilk for each group
normality_results <- dat_data %>%
  group_by(library, dataframe_size) %>%
  summarise(
    shapiro_p = shapiro.test(energy_usage)$p.value,
    .groups = "drop"
  )

normality_results
```

Interpretation: if p < 0.05, reject normality (data is not normal).

### 3.2

```{r}
bn <- bestNormalize(dat_data$energy_usage)
dat_data$norm_energy_usage <- bn$x.t
```

# 3.3

```{r}
# Histogram of normalized values
ggplot(dat_data, aes(x = norm_energy_usage)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of Normalized Energy Usage",
    x = "Normalized Energy Usage",
    y = "Frequency"
  )

# Shapiro-Wilk again for normalized values
normality_results_norm <- dat_data %>%
  group_by(library, dataframe_size) %>%
  summarise(
    shapiro_p = shapiro.test(norm_energy_usage)$p.value,
    .groups = "drop"
  )

normality_results_norm
```

### 3.4

- If most *shapiro_p* values after normalization are greater than 0.05, then distributions can be considered normal, and parametric tests (ANOVA, t-test, LMER) are justified.
- If many are still below 0.05, you should stick to non-parametric tests (Wilcoxon, ART).

## Section 4: (Non-)Parametric Tests

### 4.1

```{r, results='hide', message=FALSE}
pandas_small = dat_data[dat_data$dataframe_size == 'Small' & dat_data$library == 'Pandas',]$energy_usage
polars_small = dat_data[dat_data$dataframe_size == 'Small' & dat_data$library == 'Polars',]$energy_usage
wilcox.test(pandas_small, polars_small)

pandas_big = dat_data[dat_data$dataframe_size == 'Big' & dat_data$library == 'Pandas',]$energy_usage
polars_big = dat_data[dat_data$dataframe_size == 'Big' & dat_data$library == 'Polars',]$energy_usage
wilcox.test(pandas_big, polars_big)
```

We shall use Wilcoxon test; it's non-parametric and we can't use a parametric test as our data is not normal.

### 4.2

```{r, results='hide', message=FALSE}
library(ARTool)

model = art(energy_usage ~ library * dataframe_size + Error(trial), data = dat_data)
anova(model)
```

-   To check if model is correct we simpy run command `summary(model)` and investigate whether all F-values are 0.

-   `anova(model)` shows us information about influence of each variable and their interactions.

### 4.3

```{r, results='hide', message=FALSE}
library(lmerTest)
model = lmer(norm_energy_usage ~ library * dataframe_size + (1|trial), data = dat_data)
anova(model)
```

### 4.4

```{r}
qqnorm(resid(model))
qqline(resid(model))
shapiro.test(resid(model))
```

We can see residuals are not normal, which might be the result that our data doesn't meet the conditional normality assumption, even though it's distribution with no regard to factors looks normal.

## Section 5: Summary Plots

### 5.1 & 5.2

```{r}
p_boxplot <- ggplot(dat_data, aes(x = dataframe_size, y = energy_usage, fill = library)) +
  geom_boxplot() +
  facet_wrap(~library) +
  theme_minimal() +
  labs(
    title = "Energy Usage by Library and Dataframe Size",
    x = "Dataframe Size",
    y = "Energy Usage",
    fill = "Library"
    ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold")
    )

p_boxplot
```

### 5.3

```{r}
ggplot(dat_data, aes(x = trial, y=energy_usage, color = interaction(library,dataframe_size))) +
  geom_point(alpha = 0.75) +
  labs(
    title = "Energy Usage by Experiment Trial",
    x = "Trial",
    y = "Energy Usage",
    color = "Experiment"
  )
```

### 5.4

```{r}

ggplot(dat_data, aes(x = execution_time, y = energy_usage)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_grid(library ~ dataframe_size) +
  labs(
    title = "Energy Usage by Execution Time",
    x = "Execution Time",
    y = "Energy Usage"
  )
```

## Section 6: Annotated Tables

```{r}
# Load required libraries
library(xtable)
library(dplyr)

# Create summary statistics
summary_stats <- dat_data %>%
  group_by(library, dataframe_size) %>%
  summarise(
    cpu_avg = mean(cpu_usage),
    memory_avg = mean(memory_usage),
    energy_avg = mean(energy_usage),
    .groups = 'drop'
  )

# Create and format the xtable
summary_table <- xtable(summary_stats,
                       caption = "Summary Statistics by Library and DataFrame Size",
                       digits = c(0, 0, 0, 2, 2, 2))

# Print the table (you can adjust the type of output as needed)
print(summary_table, 
      include.rownames = TRUE,
      floating = FALSE)
```